\documentclass[12pt]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[margin=1in]{geometry}
\usepackage{parskip}
\usepackage{enumitem}
\usepackage{graphicx}
\usepackage{hyperref}  % fixed!! had to move it after graphicx, stackoverflow saved me

\title{Engineering for Clarity: Designing SimplrAI,\\
a Grade-Appropriate AI Explanation System for Student Learning}
\author{Soumil Bhandari \\ BASIS Independent McLean}
\date{April 2026}

\begin{document}
\maketitle

\section{Introduction}

Over the past two years, artificial intelligence has moved from a futuristic
concept to a core academic tool. According to the Pew Research Center (2024),
more than 63\% of American teenagers report using AI for schoolwork, and nearly
one in five uses it daily. The tools they reach for---ChatGPT, Gemini,
Perplexity---were built as general-purpose conversational agents, optimized
for breadth, fluency, and general reasoning. None were designed for the specific
cognitive needs of a thirteen-year-old trying to understand the chain rule for
the first time.

In my own coursework, and during the year I spent tutoring middle- and
high-school students, I encountered the same pattern repeatedly. A student would
ask a chatbot to explain mitosis, derivative notation, or the causes of the
French Revolution. The model would produce a fluent, dense, six-paragraph
response written somewhere between a college textbook and an encyclopedia
article. The student would skim it, decide they were still confused, and ask
for it to be simpler. The next response would often be longer than the first.

This is a design problem, not a capability problem. Frontier LLMs already have
the underlying knowledge to teach high-school content; what they lack is a
pedagogical scaffold---structured output, age-aware reading level, intentional
hallucination control, and an interface that respects what cognitive scientists
have known for forty years about working-memory limits (Sweller, 1988;
Paas \& van Merri\"enboer, 2020).

This project asks:

\begin{quote}\itshape
How can an AI-based academic explanation system be engineered to produce fast,
accurate, and grade-appropriate explanations that reduce cognitive load and
meaningfully improve student comprehension?
\end{quote}

To answer it, I designed and built SimplrAI, an explanation system for students
from sixth grade through the first years of college. SimplrAI wraps a frontier
LLM in three deliberate layers:
\begin{enumerate}[leftmargin=1.4em]
  \item A \textbf{structured prompting layer} that enforces a
  definition--example--analogy--application output schema with lightweight
  retrieval grounding against vetted curriculum sources.
  \item A \textbf{grade-level adaptation layer} that rewrites vocabulary and
  sentence complexity to match the target Lexile band.
  \item A \textbf{minimal user interface} built on progressive-disclosure
  principles that chunks explanations into expandable sections.
\end{enumerate}

Section 2 surveys the relevant literature. Section 3 describes methodology.
Sections 4--6 present results, discussion, and limitations.

\section{Literature Review}

The literature relevant to SimplrAI spans four fields: AI tutoring systems,
cognitive load theory, hallucination and factual reliability in LLMs, and
reading-level adaptation.

\subsection{AI tutoring systems}

Koedinger et al.\ (2022) found that AI tutors can support conceptual
understanding when their explanations are appropriately structured and
scaffolded. OpenAI (2023) and Anthropic (2024) both describe their models as
optimized for general helpfulness rather than pedagogical clarity. Pew Research
(2024) confirmed rapid adoption but warned that educational usefulness is limited
by inconsistent accuracy. Dwivedi et al.\ (2023) argue that LLMs produce outputs
that are syntactically fluent but epistemically unreliable.

\subsection{Cognitive load theory}

Cognitive load theory (Sweller, 1988) holds that working memory can only process
a limited amount of novel information at once. Long, dense, or disorganized
explanations increase extraneous load. Later refinements (Kalyuga, 2009; Paas
\& van Merri\"enboer, 2020) emphasize concise, sequenced, chunked explanations.
Chi and Wylie's ICAP framework (2014) finds that explanations prompting active
construction outperform passive ones. SimplrAI's output schema addresses this.

\subsection{Hallucination and factual reliability}

Maynez et al.\ (2020) document hallucination rigorously. OpenAI (2023) shows
prompting strategy influences accuracy. Two mitigation approaches: structured
prompting (Li et al., 2023) and retrieval-augmented generation (Lewis et al.,
2020). SimplrAI combines both.

\subsection{Reading-level adaptation and interface design}

The Lexile Framework (MetaMetrics, 2019) shows comprehension falls when text
exceeds a learner's threshold. Xu et al.\ (2023) show LLMs default to
college-level prose unless constrained. Nielsen Norman Group (2023) and Reich
(2020) find students learn best from interfaces with low visual noise. Quinn et
al.\ (2023) argue dense paragraphs are suboptimal for K--12 learners.

\subsection{Synthesis and gaps}

Three gaps: no deployed system targets grade-appropriate explanation; structured
prompts have not been evaluated for student comprehension; interface effects on
learning are understudied. SimplrAI addresses all three.

\section{Methodology}

\subsection{The SimplrAI system}

SimplrAI is implemented as a web application: React front end, Node/Express
backend, Supabase Postgres database, deployed at
\href{https://simplrai.tech}{\texttt{simplrai.tech}}. All completions routed
through the GPT-5 API.

The \textit{prompting layer} accepts a user question plus a target grade level
(6--14) and constructs a system prompt enforcing a fixed four-part output:
(i) a one-sentence definition, (ii) a worked example, (iii) a concrete analogy,
(iv) a short application. The prompt imposes a maximum response length and
forbids above-grade terminology without inline definition.

The \textit{retrieval layer} performs semantic search over a 4,300-document
corpus from OpenStax, Khan Academy, and CK-12; up to three retrieved passages
are passed as grounding context.

The \textit{interface layer} renders each output section as a separately
expandable card, surfaces source links, and provides a one-click ``explain it
simpler'' button.

% TODO: participants, design, procedure subsections

\end{document}
